Syntax and Semantics of Abstract Binding Trees

### 2016-01-23 ijcar-abts.tex
[ijcar] lost double struck lambda, so remove
The double struck lambda is fine, but I can't figure out how to get it
to show up here; no lambda's less confusing than a plain one.
[ijcar] update some notation
IJCAR version: import body & appendices

\usepackage{appendix}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbold}
% \usepackage{amsthm}
\usepackage{catchfilebetweentags}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{ifdraft}
\usepackage{ifthen}
\usepackage{mathtools}
\usepackage{proof}
\usepackage{scalerel}
\usepackage{setspace}
\usepackage{stmaryrd}
\usepackage{supertabular}
\usepackage{tikz-cd}
\usepackage[length=.5]{tikzpfeile}
\usepackage{todonotes}
\usepackage{url}
\usepackage[h]{esvect}

\definecolor{dark-gray}{gray}{0.20}

\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\numberwithin{equation}{section}
\renewcommand\vec[1]{\protect\vv{#1}}

\newcommand\HypCat{\mathbf{H}}
\newcommand\Supp[1]{\mathbf{supp}(#1)}
\newcommand\SupportsUnmoded[3]{#1 \blacktriangleright_{#2} #3}
\newcommand\Supports[3]{\SupportsUnmoded{\IMode{#1}}{\IMode{#2}}{\IMode{#3}}}

\newcommand\CalcFV[2]{
  \FV{\IMode{#1}}\leadsto \OMode{#2}
}

\newcommand\CalcFS[2]{
  \FS{\IMode{#1}}\leadsto \OMode{#2}
}

\newcommand\SortExp{\mathtt{exp}}

\newcommand\IsWf[4]{%
  \IMode{#1}\parallel%
  \IMode{#2}\vdash%
  \IMode{#3}\in_{\mathbf{wf}}\IMode{#4}%
}

\newcommand\Sheaves[1]{\mathbf{Sh}\left(#1\right)}
\newcommand\Presheaves[1]{\mathbf{Psh}\left(#1\right)}
\newcommand\Sheafify[1]{#1^{\#}}
\newcommand\ConstPsh[1]{\Delta[#1]}
\newcommand\ConstSh[1]{\Delta^\#[#1]}
\newcommand\SCtxSite{\SymSets[\Sorts]}
\newcommand\AtomicCoverage{J_{\mathtt{atm}}}

\DeclareMathOperator{\Perm}{\mathbf{Perm}}

@@

Abstract binding trees (abts) are a generalization of abstract syntax trees
where operators may express variable binding structure as part of their arities.
Originally formulated by Peter Aczel~\cite{aczel:1978}, unisorted abts have been
deployed successfully as the uniform syntactic framework for several
implementations of Constructive Type Theory, including
Nuprl~\cite{constable:1986}, MetaPRL~\cite{hickey:2003} and
JonPRL~\cite{jonprl:2015}.

In \emph{Practical Foundations for Programming Languages}~\cite{harper:2016},
Robert Harper develops the multi-sorted version of abstract binding trees and
proposes an extension to include families of operators indexed by
\emph{symbols}, which are, unlike variables, subject to only
distinctness-preserving renaming and not substitution; furthermore, symbols do
not appear in the syntax of abts, and are only introduced as parameters to
operators. This extension to support symbols is essential for a correct
treatment of programming languages with open sums (e.g. ML's \texttt{exn} type),
as well as assignable references.

In parallel, M.\ Fiore and his collaborators have developed the categorical
semantics for several variations of second-order algebraic
theories~\cite{fiore:1999,fiore:2005,fiore-hur:2010,fiore-mamoud:2010}; of
these, the simply sorted variants are equivalent to Harper's abstract binding
trees augmented with a notion of second-order variable (metavariable).

The contribution of this paper is the development of the syntax and semantics
of multi-sorted nominal abts, an extension of second order universal algebra to
support symbol-indexed families of operators. Additionally, we have developed
the categorical semantics for abts formally in Constructive Type Theory using
the Agda proof assistant~\cite{norell:2007}; we have also developed an
implementation for abstract binding trees in Standard ML, which we intend to
integrate into the JonPRL proof assistant \cite{jonprl:2015}.

\section{Categorical Preliminaries}

Fix a set $\Sorts$ of \emph{sorts}; we will say $\IsSort{\tau}$ when
$\Member{\tau}{\Sorts}$. In this section, we will develop a theory of sets
varying over collections of $\Sorts$-sorted symbols.

Let $\SymSets$ be the category of finite cardinals and their injective maps;
then the comma construction $\SCtx$, is the category of contexts of symbols
whose objects are finite sets of symbols $U$ and sort-assignments
\begin{tikzcd}[cramped]
  \IMode{U}\arrow[r, "\IMode{\mathfrak{s}}"] &\IMode{\Sorts}
\end{tikzcd}, and whose morphisms are
sort-preserving renamings; we will write $\Upsilon$ for a symbol context
$(U,\mathfrak{s})$.

\begin{remark}

  To be precise, $\SCtx$ is an abuse of notation for the comma category
  $\Dom{-}_\SymSets\downarrow \Delta[\Sorts]$, with $\Dom{-}_\SymSets$ the
  forgetful functor from $\SymSets$ to $\Sets$, and
  $\Define{\Delta[\Sorts](*)}{\Sorts}$ a constant functor from $\mathbb{1}$ to $\Sets$.
  Formally, an object of $\SCtx$ is, then, a triple $\Tuple{U, *,
  \mathfrak{s}}$ with $U$ an object in $\SymSets$, $*$ the unique object in
  $\mathbb{1}$, and $\mathfrak{s}$ a function $\Dom{U}_\SymSets\ra
  \Sorts$; an arrow $\Of{\varrho}{\Tuple{U,*,\mathfrak{s}}\ra
  \Tuple{V,*,\mathfrak{t}}}$ is a commuting triangle like the following:
  \[
    \begin{tikzcd}[cramped]
      \IMode{\Dom{U}_\SymSets}
        \arrow[rr, "\IMode{\Dom{\rho}_\SymSets}"]
        \arrow[dr, swap, "\IMode{\mathfrak{s}}"] &&
      \IMode{\Dom{V}_\SymSets}
      \arrow[dl, "\IMode{\mathfrak{t}}"] \\
      &\IMode{\Sorts}
    \end{tikzcd}
  \]

  Because the object $\Of{*}{\mathbb{1}}$ is unique, we will always omit it
  from the data of an object in $\SCtx$.
\end{remark}

\paragraph{Copresheaves on $\SCtx$}

A presheaf on a category $\mathcal{C}$ is a contravariant functor
$\OpCat{\mathcal{C}}\ra\Sets$; when $\Match{\mathcal{C}}{\OpCat{\mathcal{D}}}$,
we call this a \emph{copresheaf} (covariant presheaf) on $\mathcal{D}$.
Therefore, a copresheaf on $\SCtx$ is a functor $\SCtx\ra\Sets$; colloquially,
such a copresheaf can be thought of as a set that \emph{varies over} symbol
contexts.

This leads to a very intuitive setting in which to consider sets
that vary over collections of $\Sorts$-sorted symbols. In particular, for any
morphism
%
\begin{tikzcd}[cramped]
  \IMode{\Upsilon}\arrow[r, hook, "\IMode{\varrho}"] & \IMode{\Upsilon'}
\end{tikzcd}
%
in $\SCtx$, any copresheaf $\Of{X}{\Sets^\SCtx}$ has a symbol-renaming action
$\Of{X(\varrho)}{X(\Upsilon)\ra X(\Upsilon')}$. When the copresheaf $X$ is
clear from context, we will write $m\cdot\varrho$ for $X(\varrho)(m)$.

Now, when $\Member{m}{X(\Upsilon)}$ for a copresheaf $\Of{X}{\Sets^\SCtx}$, we
can define the notion that $m$ depends on \emph{at most} a subcontext
$\Upsilon'$ of $\Upsilon$; this is called \emph{support}.

\begin{definition}[Support]
  \label{def:support}
  When
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon}\arrow[r, hook, "\IMode{\varrho}"] & \IMode{\Upsilon'}
  \end{tikzcd},
  we say that $\Upsilon$ supports $\Member{m}{\Upsilon'}$ (written
  $\Supports{\Upsilon}{\varrho}{m}$) when, for all
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon'}\arrow[r, hook, "\IMode{\varrho_1}{,} \IMode{\varrho_2}"] & \IMode{\Upsilon''}
  \end{tikzcd},
  if $\IsEq{\varrho_1\circ\varrho}{\varrho_2\circ\varrho}$ then $\IsEq{m\cdot\varrho_1}{m\cdot\varrho_2}$.
\end{definition}

We can also always calculate the \emph{least support} of a term
$\Member{m}{X(\Upsilon')}$, written $\Supp{m}$:
\[
  \Define{\Supp{m}}{
    \bigcap_{
      \begin{tikzcd}[cramped, ampersand replacement = \&]
        \Upsilon\arrow[r, hook, "\varrho"] \& \Upsilon'
      \end{tikzcd}
    }
    \MkSet{
      \Upsilon
      \mid
      \SupportsUnmoded{\Upsilon}{\varrho}{m}
    }
  }
\]
Intuitively, $\Supp{m}$ is the exact symbol context that $m$ depends on.

\subsection{Sheaves on the atomic site}

When
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon}\arrow[r, hook, "\IMode{\varrho}"] & \IMode{\Upsilon'}
  \end{tikzcd}
such that $\Supports{\Upsilon}{\varrho}{m}$ for some $\Member{m}{X(\Upsilon')}$,
we would expect that we can work backward to a unique $\Of{m'}{\Upsilon}$
such that $\IsEq{m'\cdot\varrho}{m}$; whilst it is not in general the case for
presheaves $X$, this is precisely the sheaf condition for copresheaves on
$\SCtx$ under the atomic topology \cite[p.~126]{mac-lane-moerdijk:1992}, which
we will define in Definition~$\ref{def:atomic-topology}$.

The sheaf condition ensures that the notion of support is well-behaved in the
following sense: if some term lies in two different fibers, then it may be
\emph{strengthened} uniquely into the fiber of its support.

\begin{definition}[Sieves and matching families]
  A \emph{sieve} $S$ on an object $C$ is a subfunctor of the Yoneda embedding
  $\Define{\Yoneda{C}}{\Hom{\mathcal{C}}{-}{C}}$, i.e.\
  $\IMode{S}\sqsubseteq\IMode{\Yoneda{C}}$. For a presheaf $X$ on the category
  $\mathcal{C}$, a \emph{matching family} for a sieve $S$ is a natural
  transformation $\Of{\phi}{S\ra X}$. An \emph{amalgamation} for $\phi$ is an
  object $\Member{m}{X(C)}$ such that $\IsEq{m\cdot f}{\phi_D(f)}$ for each
  $\Member{f}{S(D)}$, for all objects $D$.
\end{definition}

\begin{definition}[Sites and coverages]
  A \emph{site} is a pair $\Pair{\mathcal{C}}{J}$, with $\mathcal{C}$ a category
  of ``worlds'' and $J$ a Grothendieck topology (or ``coverage''). A Grothendieck
  topology $J$ (or \emph{coverage}) is a family of sets of sieves, indexed by
  $\mathcal{C}$-objects; that is, for each object $\Of{C}{\mathcal{C}}$, $J(C)$
  is the set of $C$-sieves which will be said to $\emph{cover}$ $C$. A topology
  $J$ must satisfy several axioms (see \cite{mac-lane-moerdijk:1992} for
  details).
\end{definition}

\begin{definition}[Separation and sheafhood]
  A presheaf on a site $\Pair{\mathcal{C}}{J}$ is a presheaf on $\mathcal{C}$; we
  say that such a presheaf $X$ is \emph{separated} in case for every sieve
  $\Member{S}{J(C)}$ and matching family $\Of{\phi}{S\ra X}$, there is at most
  one amalgamation $\Member{m}{X(C)}$ for $\phi$. Additionally, $X$ is a
  \emph{sheaf} when there is \emph{exactly one} such amalgamation.
\end{definition}

\begin{definition}[The atomic topology]
  \label{def:atomic-topology}
  The atomic coverage $\AtomicCoverage$ is the one in which all non-empty sieves
  cover; in other words:
  \[
    \Define{
      \AtomicCoverage(C)
    }{
      \MkSet{
        S\sqsubseteq\Yoneda{C}
        \mid
        \textstyle
        \bigcup_D S(D)\ \textit{non-empty}
      }
    }
  \]
\end{definition}

In the future, we will write $\SCtxSite$ for the Grothendieck site
$\Pair{\OpCat{\left(\SCtx\right)}}{\AtomicCoverage}$.
Sheafhood on the atomic site $\SCtxSite$ amounts to the
following condition, using Definition~\ref{def:support}:
\begin{quote}
  A copresheaf $\Of{X}{\Sets^\SCtx}$ is a sheaf on $\SCtxSite$ just when, for all $\Member{n}{X(\Upsilon')}$, if we have
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon}\arrow[r, hook, "\IMode{\varrho}"] & \IMode{\Upsilon'}
  \end{tikzcd}
  and $\Supports{\Upsilon}{\varrho}{n}$, then there is a unique
  $\Member{m}{X(\Upsilon)}$ such that $\IsEq{m\cdot\varrho}{n}$.
\end{quote}

Therefore, a sheaf on $\SCtxSite$ is a set that varies over $\Sorts$-sorted
symbol contexts, which has a well-behaved notion of support. Another way to put
it is, if $\Member{m}{X(\Upsilon)}$ only ``uses'' some of the symbols available
at world $\Upsilon$, there's a canonical way to restrict $m$ to a term in the
world which contains \emph{only} the symbols that it uses.

\paragraph{Constructions on sheaves}

%% For any Grothendieck site $\Match{\mathfrak{S}}{\Pair{\mathcal{C}}{J}}$, the
%% inclusion $\Of{\iota}{\Sheaves{\mathfrak{S}}\ra\Presheaves{\mathcal{C}}}$ of
%% sheaves into presheaves has a left adjoint
%% $\IMode{\Sheafify{(-)}}\dashv\IMode{\iota}$, which takes any presheaf
%% $\Of{X}{\Presheaves{\mathcal{C}}}$ to a unique sheaf
%% $\Of{\Sheafify{X}}{\Sheaves{\mathfrak{S}}}$, called the \emph{sheafification}
%% of $X$. Sheafification is the twice-iterated application of the
%% \emph{+-construction} $X^+$, which turns a presheaf into a separated presheaf,
%% and a separated presheaf into a sheaf \cite{mac-lane-moerdijk:1992}:
%% \begin{align*}
%%   \ADefine{
%%     X^+(C)
%%   }{
%%     \int^{S\in J(C)}
%%     \Hom{\Presheaves{\mathcal{C}}}{S}{X}
%%   }\\
%%   \ADefine{
%%     \Sheafify{X}
%%   }{
%%     X^{++}
%%   }
%% \end{align*}
%%
%% We also have a canonical natural transformation $\Of{\eta}{X\ra X^+}$, defined
%% using the maximal sieve $\Define{t_C}{\Yoneda{C}}$:
%% \[
%%   \HypJ{
%%     \Define{\eta_C(m)}{
%%       \Pair{t_C}{
%%         X(-)(m)
%%       }
%%     }
%%   }{
%%     \Member{m}{X(C)}
%%   }
%% \]

Every category of sheaves on a site gives rise to a topos, which equips us with
a number of standard constructions, including (among other things) the disjoint
union of sheaves $X\oplus Y$, the product of sheaves $X\otimes Y$, and the terminal sheaf $\mathbb{1}$.

\section{Operators and signatures}

A valence $\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}$ specifies an expression
of sort $\tau$ which binds symbols in $\vec{\sigma}$ and variables in
$\vec{\tau}$.

\[
  \infer{
    \IsValence{
      \MkValence
        {\sigma_0,\dots,\sigma_m}
        {\tau_0,\dots,\tau_n}
        {\tau}
    }
  }{
    \IsSort{\tau}
&
    \HypJ{\IsSort{\sigma_i}}{\Leq{i}{m}}
&
    \HypJ{\IsSort{\tau_i}}{\Leq{i}{n}}
  }
\]

An arity $\MkArity{\vec{v}}{\tau}$ specifies an operator of sort $\tau$ with
arguments of valences $\vec{v}$. We will call the set of valences $\Valences$,
and the set of arities $\Arities$.

\[
  \infer{
    \IsArity{
      \MkArity
        {v_0,\ldots,v_n}
        {\tau}
    }
  }{
    \IsSort{\tau}
&
    \HypJ
      {\IsValence{v_i}}
      {\Leq{i}{n}}
  }
\]

Fix a family of sheaves of operators
$\Member{\Operators_a}{\Sheaves{\SCtxSite}}$, indexed by arities
$\Member{a}{\Arities}$; for each $\Operators_a$, arrows in $\SCtx$ will lift to
renamings in operators' symbolic parameters. Together, $\Sorts$ and $\Operators$ are said to form a \emph{signature}
$\Define{\Sigma}{\Pair{\Sorts}{\Operators}}$.

We will write $\IsOperator{\Upsilon}{\vartheta}{a}$ in case
$\Member{\vartheta}{\Operators_a(\Upsilon)}$; note that this judgment enjoys
the structural properties of weakening and exchange via functoriality of
$\Operators$, and strengthening via the sheaf condition. An operator signature
is defined by specifying, for each arity $a$, the sheaf $\Operators_a$, whose
``elements'' are the operators of arity $a$.

\paragraph{Examples}

For instance, consider a $\lambda$-calculus with a single sort, $\SortExp$; we
give its signature $\Sigma_{\lambda}$ by asserting the following about its
operators:
\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{lam}}{
    \MkArity{
      \MkValence{}{\SortExp}{\SortExp}
    }{\SortExp}
  }\\
  \IsOperator{\Upsilon}{\mathtt{fix}}{
    \MkArity{
      \MkValence{}{\SortExp}{\SortExp}
    }{\SortExp}
  }\\
  \IsOperator{\Upsilon}{\mathtt{ap}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{}{}{\SortExp}
    }{\SortExp}
  }
\end{gather*}

These rules correspond to the following definition of $\Operators$:
\begin{align*}
  \ADefine{
    \Operators_{
       \MkArity{
        \MkValence{}{\SortExp}{\SortExp}
      }{\SortExp}
    }
  }{
    \mathbb{1}\oplus\mathbb{1}
  }
  \\
  %
  \ADefine{
    \Operators_{
      \MkArity{
        \MkValence{}{}{\SortExp},
        \MkValence{}{}{\SortExp}
      }{\SortExp}
    }
  }{
    \mathbb{1}
  }
\end{align*}

So far, we have made no use of symbols and parameters; however, consider the
extension of the calculus with assignables (references):
%
\begin{gather*}
  \IsOperator{\Upsilon}{\mathtt{decl}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{\SortExp}{}{\SortExp}
    }{
      \SortExp
    }
  }
\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{get}[u]}{
    \MkArity{}{\SortExp}
  }
\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathtt{set}[u]}{
    \MkArity{
      \MkValence{}{}{\SortExp}
    }{
      \SortExp
    }
  }
\end{gather*}

Let $\mathbf{S}_\tau$ be the sheaf of symbols of sort $\tau$, a subobject of
the Yoneda embedding of the empty symbol context, $\Yoneda{\cdot}$:
\[
  \Define{\mathbf{S}_\tau(\Upsilon)}{
    \MkSet{
      u\in\Dom{\Upsilon}
      \mid
      \Upsilon(u) \equiv \tau
    }
  }
\]
%
Then the above rules correspond to the following family of sheaves:
\begin{align*}
  \ADefine{
    \Operators_{
      \MkArity{
        \MkValence{}{}{\SortExp},
        \MkValence{\SortExp}{}{\SortExp}
      }{
        \SortExp
      }
    }
  }{
    \mathbb{1}
  }\\
  \ADefine{
    \Operators_{
      \MkArity{}{\SortExp}
    }
  }{
    \mathbf{S}_\SortExp
  }\\
  \ADefine{
    \Operators_{
      \MkArity{
        \MkValence{}{}{\SortExp}
      }{
        \SortExp
      }
    }
  }{
    \mathbf{S}_\SortExp
  }
\end{align*}

Declaring a new assignable consists in providing an initial value, and an
expression binding a symbol (which shall represent the assignable in scope).
Note that the functoriality of $\Operators$ guarantees for any renaming
$\Of{\varrho}{\Upsilon\inj\Upsilon'}$, a family of lifted maps
$\Of{\Operators_a(\varrho)}{\Operators_a(\Upsilon)\inj\Operators_a(\Upsilon')}$
natural in $a$, such that
$\IsOperator{\Upsilon'}{\Operators(\varrho)(\vartheta)}{a}$ when
$\IsOperator{\Upsilon}{\vartheta}{a}$. In particular, the renaming
$\Upsilon,u\mapsto\Upsilon,v$ shall take $\mathtt{get}[u]$ to
$\mathtt{get}[v]$. In the future, we will write $\Rename{\varrho}{\vartheta}$
for $\Operators_a(\varrho)(\vartheta)$.

\section{Contexts}

In general, we will have three kinds of context: symbolic (parameter) contexts,
variable contexts, and metavariable contexts. The symbol contexts have already
been defined via the comma construction $\SCtx$, but they also admit a
syntactic characterization,
\begin{gather*}
  \infer{
    \IsSymCtx{\cdot}
  }{
  }
\qquad
  \infer{
    \IsSymCtx{\Upsilon,u:\tau}
  }{
    \IsSymCtx{\Upsilon}
&
    \IsSort{\tau}
&
    \NotIn{u}{\Dom\Upsilon}
  }
\end{gather*}
%
Because, modulo notation, we have $\Member{\Upsilon}{\SCtx}$ just when
$\IsSymCtx{\Upsilon}$, we will use the syntactic view when it is convenient.

Contexts of variables are similar to contexts of symbols, except that they
admit \emph{any} renamings, not just the injective ones. As such, when
$\FinSets$ is the category of finite cardinals and all functions between them, the
comma construction $\Ctx$ is the category of variable contexts. As above, we
can give them an equivalent syntactic treatment:
\begin{gather*}
  \infer{
    \IsVarCtx{\cdot}
  }{
  }
\qquad
  \infer{
    \IsVarCtx{\Gamma,x:\tau}
  }{
    \IsVarCtx{\Gamma}
&
    \IsSort{\tau}
&
    \NotIn{x}{\Dom\Gamma}
  }
\end{gather*}

A metavariable context consists of bindings of $\Sorts$-\emph{valences} to
metavariables; let $\Define{\Valences}{\MkSet{v\mid \IsValenceUnmoded{v}}}$ be the set
of valences. Then, the category of metavariable contexts is the comma
construction $\MCtx$, which likewise admits an equivalent inductive definition:
\begin{gather*}
  \infer{
    \IsMetaCtx{\cdot}
  }{
  }
\qquad
  \infer{
    \IsMetaCtx{\Theta,\MV{m}:v}
  }{
    \IsMetaCtx{\Theta}
&
    \IsValence{v}
&
    \NotIn{\MV{m}}{\Dom\Theta}
  }
\end{gather*}

\section{Nominal Abstract Binding Trees}

Let the judgment $\IsAbt{\Theta}{\Upsilon}{\Gamma}{M}{\tau}$ presuppose
\footnote{
  In the \emph{judgmental method}, as pioneered by Per Martin-L\"of, a form of
  judgment is propounded first by specifying its range of significance (i.e.\
  the circumstances under which it shall have a meaning), and then, a meaning
  explanation (definition) for the judgment is given. The range of significance
  of a judgment is called its \emph{presupposition}, and the meaning of the
  judgment may proceed by induction on the evidence that the presupposition
  obtains. See \cite{schroeder-heister:1987} for a more detailed explanation of
  presuppositions and the judgmental method.
}
$\IsMetaCtx{\Theta}$, $\IsSymCtx\Upsilon$, $\IsVarCtx\Gamma$ and
$\IsSort{\tau}$, meaning that $M$ is an abstract binding tree of sort $s$, with
metavariables in $\Theta$, parameters in $\Upsilon$, and variables in $\Gamma$.
Let the judgment $\IsAbs{\Theta}{\Upsilon}{\Gamma}{E}{v}$ presuppose
$\IsValence{v}$. Then, the syntax of abstract binding trees is inductively
defined in four rules:

\[
  \infer[\RuleIsAbtVar]{
    \IsAbt{\Theta}{\Upsilon}{\Gamma}{x}{\tau}
  }{
    \Lookup{\Gamma}{x}{\tau}
  }
\]
\[
  \infer[\RuleIsAbtMVar]{
    \IsAbt{\Theta}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{u_0,\dots,u_m}{M_0,\dots,M_n}
    }{\tau}
  }{
    \begin{array}{l}
      \Lookup{\Theta}{\MV{m}}{
        \MkValence{
          \sigma_0,\dots,\sigma_m
        }{
          \tau_0,\dots,\tau_n
        }{\tau}
      }
\\
      \HypJ{
        \Lookup{\Upsilon}{u_i}{\sigma_i}
      }{\Leq{i}{m}}
\\
      \HypJ{
        \IsAbt{\Theta}{\Upsilon}{\Gamma}{M_i}{\tau_i}
      }{\Leq{i}{n}}
    \end{array}
  }
\]
\[
  \infer[\RuleIsAbtApp]{
    \IsAbt{\Theta}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{\tau}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        \MkArity{v_1,\dots,v_n}{\tau}
      }
\\
      \HypJ{
        \IsAbs{\Theta}{\Upsilon}{\Gamma}{E_i}{v_i}
      }{\Leq{i}{n}}
    \end{array}
  }
\]
\[
  \infer[\RuleIsAbtAbs]{
    \IsAbs{\Theta}{\Upsilon}{\Gamma}{
      \MkAbs{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}
    }
  }{
    \IsAbt{
      \Theta
    }{
      \Upsilon, \vec{u}:\vec{\sigma}
    }{
      \Gamma,\vec{x}:{\vec{\tau}}
    }{
      M
    }{\tau}
  }
\]

Abts are identified up to $\alpha$-equivalence.

\subsection{Calculating free variables}

\newcommand\RuleFVVar{\RuleNameVar{\mathbf{FV}}{}}
\newcommand\RuleFVMVar{\RuleNameMVar{\mathbf{FV}}{}}
\newcommand\RuleFVApp{\RuleNameApp{\mathbf{FV}}}
\newcommand\RuleFVAbs{\RuleNameAbs{\mathbf{FV}}{}}

We can easily calculate the variables free in an expression by recursion on its
structure:

\begin{gather*}
  \infer[\RuleFVVar]{
    \CalcFV{x}{\MkSet{x}}
  }{
  }\\[6pt]
  %
  \infer[\RuleFVMVar]{
    \CalcFV{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \bigcup_{i\leq n} \vec{x}_i
    }
  }{
    \HypJ{
      \CalcFV{M_i}{\vec{x}_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  %
  \infer[\RuleFVApp]{
    \CalcFV{
      \App{\vartheta}{E_0,\dots,E_n}
    }{
      \bigcup_{i\leq n} \vec{x}_i
    }
  }{
    \HypJ{
      \CalcFV{E_i}{\vec{x}_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  %
  \infer[\RuleFVAbs]{
    \CalcFV{
      \MkAbs{\vec{u}}{\vec{y}}{M}
    }{
      \vec{x}\setminus\vec{y}
    }
  }{
    \CalcFV{M}{\vec{x}}
  }
\end{gather*}

Because this is a total relation, henceforth we will write $\FV{M}$ for
$\vec{x}$ when $\CalcFV{M}{\vec{x}}$.

\subsection{Calculating free symbols}

\newcommand\RuleFSVar{\RuleNameVar{\mathbf{FS}}{}}
\newcommand\RuleFSMVar{\RuleNameMVar{\mathbf{FS}}{}}
\newcommand\RuleFSApp{\RuleNameApp{\mathbf{FS}}}
\newcommand\RuleFSAbs{\RuleNameAbs{\mathbf{FS}}{}}

Whereas the calculation of free variables pivoted on the $\RuleFVVar$ rule, the
calculation of free symbols will pivot on the $\RuleFSApp$ and $\RuleFSMVar$ rules,
because the only way a symbol can be introduced is as a parameter to an
operator or as a parameter to a metavariable.

\begin{gather*}
  \infer[\RuleFSVar]{
    \CalcFS{x}{\MkSet{}}
  }{
  }\\[6pt]
  %
  \infer[\RuleFSMVar]{
    \CalcFS{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \vec{u} \cup \bigcup_{i\leq n} \vec{u}_i
    }
  }{
    \HypJ{
      \CalcFS{M_i}{\vec{u}_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  %
  \infer[\RuleFSApp]{
    \CalcFS{
      \App{\vartheta}{E_0,\dots,E_n}
    }{
      \Dom{\Supp{\vartheta}} \cup \bigcup_{i\leq n} \vec{u}_i
    }
  }{
    \HypJ{
      \CalcFS{E_i}{\vec{u}_i}
    }{
      \Leq{i}{n}
    }
  }\\[6pt]
  %
  \infer[\RuleFSAbs]{
    \CalcFS{
      \MkAbs{\vec{v}}{\vec{x}}{M}
    }{
      \vec{u}\setminus\vec{v}
    }
  }{
    \CalcFS{M}{\vec{u}}
  }
\end{gather*}

Because this is a total relation, henceforth we will write $\FS{M}$ for
$\vec{u}$ when $\CalcFS{M}{\vec{u}}$.

\subsection{Renaming of symbols}

\newcommand\RuleRenPrefix{\bullet}
\newcommand\RuleRenVar{\RuleNameVar{\RuleRenPrefix}{}}
\newcommand\RuleRenMVar{\RuleNameMVar{\RuleRenPrefix}{}}
\newcommand\RuleRenApp{\RuleNameApp{\RuleRenPrefix}}
\newcommand\RuleRenAbs{\RuleNameAbs{\RuleRenPrefix}{}}

The only place that symbols appear in our calculus is as parameters to operators
(unlike variables, symbols are not terms). Therefore, the functorial action of
the operator sheaf can be lifted into terms by recursion on their
structure, via a pair of judgments $\IsRenaming{\varrho}{M}{N}$ and
$\IsRenaming{\varrho}{E}{F}$, presupposing $\Of{\varrho}{\Upsilon\inj\Upsilon'}$,
and $\IsAbt{\Theta}{\Upsilon}{\Gamma}{M}{\tau}$ and
$\IsAbt{\Theta}{\Upsilon}{\Gamma}{E}{v}$ respectively:

\begin{gather*}
  \infer[\RuleRenVar]{
    \IsRenaming{\varrho}{x}{x}
  }{
  }
\\[6pt]
  \infer[\RuleRenMVar]{
    \IsRenaming{\varrho}{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{
      \MApp{\MV{m}}{\vec{v}}{N_0,\dots,N_n}
    }
  }{
    \Match{\varrho(\vec{u})}{\vec{v}}
&
    \HypJ{
      \IsRenaming{\varrho}{M_i}{N_i}
    }{
      \Leq{i}{n}
    }
  }
\\[6pt]
  \infer[\RuleRenApp]{
    \IsRenaming{\varrho}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{
      \App{\Rename{\varrho}{\vartheta}}{F_0,\dots,F_n}
    }
  }{
    \HypJ{
      \IsRenaming{\varrho}{E_i}{F_i}
    }{
      \Leq{i}{n}
    }
  }
\\[6pt]
  \infer[\RuleRenAbs]{
    \IsRenaming{\varrho}{
      \MkAbs{\vec{u}}{\vec{x}}{M}
    }{
      \MkAbs{\vec{u}}{\vec{x}}{N}
    }
  }{
    \IsRenaming{\varrho\setminus\vec{u}}{M}{N}
  }
\end{gather*}

Above, the notation $\varrho\setminus\vec{u}$ means the omission of the
variables $\vec{u}$ from the renaming $\varrho$. Because terms are identified up
to $\alpha$-equivalence, the renaming judgment is functional in its input, and
so we are justified in writing $\Rename{\varrho}{M}$ for $N$ when
$\IsRenaming{\varrho}{M}{N}$.

\subsection{Substitution of variables}

\newcommand\RuleSubstPrefix{/}
\newcommand\RuleSubstVar[1]{\RuleNameVar{\RuleSubstPrefix}{#1}}
\newcommand\RuleSubstMVar{\RuleNameMVar{\RuleSubstPrefix}{}}
\newcommand\RuleSubstApp{\RuleNameApp{\RuleSubstPrefix}}
\newcommand\RuleSubstAbs[1]{\RuleNameAbs{\RuleSubstPrefix}{#1}}

Variable substitution in abts is defined inductively by a pair of judgments,
$\IsSubst{N}{x}{M}{M'}$ and $\IsSubst{N}{x}{E}{F}$:

>>>>

\usepackage{appendix}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbold}
% \usepackage{amsthm}
\usepackage{catchfilebetweentags}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{ifdraft}
\usepackage{ifthen}
\usepackage{mathtools}
\usepackage{proof}
\usepackage{scalerel}
\usepackage{setspace}
\usepackage{stmaryrd}
\usepackage{supertabular}
\usepackage{tikz-cd}
\usepackage[length=.5]{tikzpfeile}
\usepackage{todonotes}
\usepackage{url}
\usepackage[h]{esvect}
\usepackage{ssabt}

\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\numberwithin{equation}{section}

\renewcommand\vec[1]{\protect\vv{#1}}
\renewcommand\MkAbs[3]{\{#1\}[#2].\, #3}

@@

\input{body}

\appendix
\appendixpage
\input{appendices}
